### 2022-08-10 manuscript.tex
Merge pull request #271 from SORTEE-Github-Hackathon/RCO-last-edits [ci skip]
This build is based on
https://github.com/SORTEE-Github-Hackathon/manuscript/commit/b2d3efb7b1c0fed0477b041fbe9cb7dbc487d31c.

This commit was created by the following CI build and job:
https://github.com/SORTEE-Github-Hackathon/manuscript/commit/b2d3efb7b1c0fed0477b041fbe9cb7dbc487d31c/checks
https://github.com/SORTEE-Github-Hackathon/manuscript/actions/runs/2834762391

This manuscript
(\href{https://SORTEE-Github-Hackathon.github.io/manuscript/v/603d35edf8325ad77fc38594ae2a81ef11346da3/}{permalink})
was automatically generated
from \href{https://github.com/SORTEE-Github-Hackathon/manuscript/tree/603d35edf8325ad77fc38594ae2a81ef11346da3}{SORTEE-Github-Hackathon/manuscript@603d35e}
on August 10, 2022.

@@

With over 83 million registered users as of 2022, GitHub is the most widely-used web platform for collaborating on computer code\textsuperscript{\protect\hyperlink{ref-nwCtHDCn}{6}}.
GitHub provides a simple but powerful web interface that allows users to participate in projects by contributing, modifying and discussing existing code, reporting bugs, discovering code and data, and publishing new code.
Through version control, users have a detailed, chronological record of the files and directories stored in their repositories\textsuperscript{\protect\hyperlink{ref-RVetqmsg}{7}} (see \protect\hyperlink{definitions}{Box 1}).
This workflow provides a strong and clear advantage over sending files back-and-forth (\emph{e.g.}, via email), a process that can become challenging and time-consuming in long-term and collaborative projects\textsuperscript{\protect\hyperlink{ref-4ny1onB0}{8}}.
Through its combination of version control and collaborative features, GitHub facilitates open source code alongside collaborative development\textsuperscript{\protect\hyperlink{ref-kEX5dgzK}{9}}.

@@

\textbf{commit}: Commits are snapshots of the development of a project. In Git, versions of files and directories are uniquely identified as ``commits'', allowing one to identify and track modifications line-by-line. Commits can include changes in multiple files and must include a brief commit message describing the changes made. A typical workflow is to make some related changes in files, add a commit message (\emph{e.g.} ``Generate and include results figure''), and after several commits \textbf{push} those commits to the remote (\emph{i.e.}, cloud-based) GitHub \textbf{repository}. \\

@@

Automation has a strong potential to expand the scale and pace of research in ecology and evolution\textsuperscript{\protect\hyperlink{ref-cfgXxgt1}{54}}.
Automation frameworks can streamline many stages of the scientific process, including automated data collection and data validation\textsuperscript{\protect\hyperlink{ref-lXvpQxeN}{56}}, automated data analysis\textsuperscript{\emph{e.g.}, \protect\hyperlink{ref-Qh7xTLwz}{57}}, automated archiving and deployment of data, code and reports\textsuperscript{\emph{e.g.}, this manuscript, \protect\hyperlink{ref-SirQKFIz}{58}}.
In this context, small modifications to code and data can be frequently committed and automatically tested, as in continuous integration and continuous deployment practices\textsuperscript{\protect\hyperlink{ref-1Ep9EJL6y}{59}}.
This allows for early detection and correction of errors, potentially improving confidence in scientific development by minimizing software errors\textsuperscript{see \protect\hyperlink{ref-ufw0ZdnI}{60}}.
In addition to increasing scientific rigor and confidence in ecological software\textsuperscript{\protect\hyperlink{ref-12103x16N}{61}}, automation can help advance more rapidly sharing ecological data and making sure the data are high quality\textsuperscript{\protect\hyperlink{ref-rTbinQMj}{62}}.
Integrating automation workflows has been highly encouraged in areas of EEB, including predictive ecology\textsuperscript{\protect\hyperlink{ref-mmCOSRfr}{63}}, long-term ecological studies\textsuperscript{\protect\hyperlink{ref-1CJo8lo2v}{64},\protect\hyperlink{ref-RxK4CmfR}{65}}, and management of species at risk information\textsuperscript{\protect\hyperlink{ref-kZzfmBNu}{66}}.

Automation can be integrated into GitHub repositories through the GitHub Actions feature\textsuperscript{\protect\hyperlink{ref-NUXbp429}{67}}, or through alternative automation systems (\emph{e.g.}, Circle CI\textsuperscript{\protect\hyperlink{ref-T03Api6e}{68}}, Travis CI\textsuperscript{\protect\hyperlink{ref-YeFSbfFV}{69}}).
Users can set up workflows associated with their repositories that are triggered by events (\emph{e.g.}, push, pull request or at specified times) for remote servers to perform user-specified steps and actions.
These actions are highly configurable and have numerous applications, such as automatically running analyses and creating figures when data or code are updated, incorporating changes to websites or applications, testing modifications to software (\emph{e.g.}, R or Python libraries).
Action workflows can be found in GitHub's Marketplace (https://github.com/marketplace?type=actions) or, alternatively, in open user repositories.

@@

As an example, GitHub Organizations are particularly well-suited to host documents and projects within a laboratory, such as research compendia, codes of conduct, protocols, training documents, and other relevant documents that evolve collaboratively over time.
In this way, teams have full ownership of repositories within an organization, while ensuring that these materials stay accessible to the laboratory after people have moved on or when locally-stored data are lost.
This application extends to research centres, which may include several distinct projects that remain linked to institutions {[}\emph{e.g.}, the German Centre for Integrative Biodiversity Research (iDiv, \url{https://github.com/idiv-biodiversity}){]}.
The team organizing the hackathon which inspired this article used a GitHub Organization (SORTEE-Github-Hackathon, \url{https://github.com/SORTEE-Github-Hackathon}) to centralize the project development, from meeting notes to, ultimately, this manuscript.
Organizations are also convenient for hosting learning materials, including lectures or workshops, such as the Québec Centre for Biodiversity Science R Workshop Series (QCBSRworkshops, \url{https://github.com/QCBSRworkshops}) or the University of Edinburgh's Coding Club (Coding Club, \url{https://github.com/ourcodingclub}), which may be continuously updated by an ever-evolving group of contributors.

@@

There have been many calls for researchers outside of the software development community to adopt GitHub to improve their collaborative research\textsuperscript{\protect\hyperlink{ref-10ghgV3S8}{3},\protect\hyperlink{ref-1Du6fzB8g}{39},\protect\hyperlink{ref-UsTxAq4f}{70}}.
This call comes in light of the continuous shift towards open-science and the increasing computational and data requirements in EEB.
Until now, resources and practical guidance on using GitHub within the EEB community have been dispersed in blog posts and video tutorials (\protect\hyperlink{tips}{Box 2}).
These resources have been useful for learning to use GitHub in our own work.
We expect that situating the main uses of GitHub in EEB into one medium, while adding on our personal perspectives, will be useful to the EEB community.

@@

First, real-time document editing is still best performed on other platforms (\emph{e.g.}, cloud-stored documents from Microsoft Word, Google Docs, hackMD (\url{https://hackmd.io/})).
Second, operations that are dependent on other software might not be easily achievable in GitHub, such as manipulating figures or tables.
Although creating tables and figures can be done through code, users may choose other software to collaboratively brainstorm figures and tables (\emph{e.g}. Google Slides, Google Sheets; but see GitHub Discussions).

@@

Lastly, fourth, GitHub may increase the distinctions between free and paid-for plans.
When projects become highly collaborative, they may have to pay GitHub features, such for branch protections, multiple reviewers of pull requests and time in its automation tools.
Currently, GitHub offers Education Packs (https://education.github.com/) to students and academics, which extends some paid features to the free plan.
However, the acquisition of GitHub by Microsoft has raised concerns over the future of free plans, causing several biodiversity data managers to shift to alternative Open Source Git services (\emph{e.g.}, Bitbucket and GitLab).

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 0\tabcolsep) * \real{1.0000}}@{}}
\toprule
\begin{minipage}[b]{\linewidth}\raggedright
Ten tips for getting started in GitHub
\end{minipage} \\
\midrule
\endhead
1. \textbf{Check for existing solutions to your problem.} The GitHub Help webpage (\url{https://docs.github.com/en}) contains extensive and detailed documents with helpful screenshots. It is a good starting point for handling an issue, and has troubleshooting tips for specific problems. Alternatively, consider Tweeting your issue. There is a large community of GitHub users around the world who have likely faced analogous problems and may be able to provide quick solutions. Third, try to follow blogs (\emph{e.g.}, \url{https://github.blog}), Twitter accounts or YouTube channels that regularly post practical solutions about the most widely-used web platform for common GitHub issues. \\
2. \textbf{Consider taking free courses}, such as those from Software Carpentry\textsuperscript{\protect\hyperlink{ref-pjy75gHr}{73}}, and sharing these courses with your lab members or colleagues. \\
3. \textbf{Take advantage of GitHub as an asynchronous working tool for team-based projects.} See the repository for this paper (\url{https://github.com/SORTEE-Github-Hackathon/manuscript}) as an example of a collaboratively authored manuscript that used the GitHub Discussions, Issues, Pages, and Actions features. \\
4. \textbf{Use the interactive courses from the GitHub Skills page (\url{https://skills.github.com/})}, which allow you to learn GitHub basics through short projects and tasks with step-by-step guides. \\
5. \textbf{Learn markdown and use cheatsheets (\emph{e.g.}, \url{http://markdownguide.org/basic-syntax})} so you can write clear metadata README files for your repositories. \\
6. \textbf{Consult online resources.} The Jenny Bryan Universe of GitHub material, for example, provides a thorough and accessible introduction for a multitude of research-related uses for GitHub, and includes a book\textsuperscript{\protect\hyperlink{ref-ZvrOcg9w}{74}}, statistics course\textsuperscript{\protect\hyperlink{ref-6CMMeSeD}{75}} and academic article\textsuperscript{\protect\hyperlink{ref-RVetqmsg}{7}}. \\
7. \textbf{Do not be afraid of trial-and-error.} One of the best ways to learn GitHub is the ``trial-and-error'' method. Learning from your own mistakes can be a valuable way to master your GitHub abilities. In any case, if you make mistakes, GitHub allows you to revert any steps that you desire via version controlling. \\
8. \textbf{If you are an educator, include lectures on reproducibility and tools for creating reproducible workflows in the curricula.} Some graduate programs include coursework on course R Markdown and GitHub. Getting students started with these tools earlier will prevent the resistance that comes from working with a less reproducible workflow for a longer period of time. (see example \url{https://github.com/rmcelreath/stat_rethinking_2022}) \\
9. \textbf{Try to begin committing with graphical user interfaces (GUI) instead of command line interfaces (CLI).} Examples of GUI are the GitHub Desktop (\url{https://desktop.github.com}), git-gui (\url{https://git-scm.com/docs/git-gui}), RStudio (\url{https://www.rstudio.com}), Visual Studio Code (\url{https://code.visualstudio.com}), Atom (\url{https://atom.io}), GitKraken (\url{https://www.gitkraken.com}). \\
10. \textbf{Get help deciphering GitHub Notifications.} Try using tools like Octobox (\url{https://octobox.io}) to disentangle and manage multiple notifications from distinct GitHub projects. \\
\bottomrule
\end{longtable}

@@

>>>>

This manuscript
(\href{https://SORTEE-Github-Hackathon.github.io/manuscript/v/b2d3efb7b1c0fed0477b041fbe9cb7dbc487d31c/}{permalink})
was automatically generated
from \href{https://github.com/SORTEE-Github-Hackathon/manuscript/tree/b2d3efb7b1c0fed0477b041fbe9cb7dbc487d31c}{SORTEE-Github-Hackathon/manuscript@b2d3efb}
on August 10, 2022.

@@

With over 83 million registered users as of 2022, GitHub is the most widely-used web platform for collaborating on computer code\textsuperscript{\protect\hyperlink{ref-nwCtHDCn}{6}}.
GitHub provides a simple but powerful web interface that allows users to participate in projects by contributing, modifying and discussing existing code, reporting bugs, discovering code and data, and publishing new code.
Through version control, users have a detailed, chronological record of the files and directories stored in their repositories\textsuperscript{\protect\hyperlink{ref-RVetqmsg}{7}} (see \protect\hyperlink{definitions}{Box 1}).
This workflow provides a strong and clear advantage over sending files back-and-forth (\emph{e.g.}, via email), a process that can become challenging and time-consuming in long-term and collaborative projects\textsuperscript{\protect\hyperlink{ref-4ny1onB0}{8}}.
Through its combination of version control and collaborative features, GitHub facilitates open-source code alongside collaborative development\textsuperscript{\protect\hyperlink{ref-kEX5dgzK}{9}}.

@@

\textbf{commit}: Commits are snapshots of the development of a project. In Git, versions of files and directories are uniquely identified as ``commits'', allowing one to identify and track modifications line-by-line. Commits can include changes in multiple files and must include a brief commit message describing the changes made. A typical workflow is to make some related changes in files, add a commit message (\emph{e.g.}, ``Generate and include results figure''), and after several commits \textbf{push} those commits to the remote (\emph{i.e.}, cloud-based) GitHub \textbf{repository}. \\

@@

Automation has a strong potential to expand the scale and pace of research in ecology and evolution\textsuperscript{\protect\hyperlink{ref-cfgXxgt1}{54}}.
Automation frameworks can streamline many stages of the scientific process, including automated data collection and data validation (\emph{e.g.},\textsuperscript{\protect\hyperlink{ref-bhDgD6lF}{55},\protect\hyperlink{ref-lXvpQxeN}{56}}), automated data analysis (\emph{e.g.},\textsuperscript{\protect\hyperlink{ref-Qh7xTLwz}{57}}), automated archiving and deployment of data, code and reports (\emph{e.g.}, this manuscript,\textsuperscript{\protect\hyperlink{ref-SirQKFIz}{58}}).
In this context, small modifications to code and data can be frequently committed and automatically tested, as in continuous integration and continuous deployment practices\textsuperscript{\protect\hyperlink{ref-1Ep9EJL6y}{59}}.
This allows for early detection and correction of errors, potentially improving confidence in scientific development by minimizing software errors (see\textsuperscript{\protect\hyperlink{ref-ufw0ZdnI}{60}}).
In addition to increasing scientific rigor and confidence in ecological software\textsuperscript{\protect\hyperlink{ref-12103x16N}{61}}, automation can help advance more rapidly sharing ecological data and making sure the data are high quality\textsuperscript{\protect\hyperlink{ref-rTbinQMj}{62}}.
Integrating automation workflows has been highly encouraged in areas of EEB, including predictive ecology\textsuperscript{\protect\hyperlink{ref-mmCOSRfr}{63}}, long-term ecological studies\textsuperscript{\protect\hyperlink{ref-1CJo8lo2v}{64},\protect\hyperlink{ref-RxK4CmfR}{65}}, and management of species at risk information\textsuperscript{\protect\hyperlink{ref-kZzfmBNu}{66}}.

Automation can be integrated into GitHub repositories through the GitHub Actions feature\textsuperscript{\protect\hyperlink{ref-NUXbp429}{67}}, or through alternative automation systems (\emph{e.g.}, Circle CI\textsuperscript{\protect\hyperlink{ref-T03Api6e}{68}}, Travis CI\textsuperscript{\protect\hyperlink{ref-YeFSbfFV}{69}}).
Users can set up workflows associated with their repositories that are triggered by events (\emph{e.g.}, push, pull request or at specified times) for remote servers to perform user-specified steps and actions.
These actions are highly configurable and have numerous applications, such as automatically running analyses and creating figures when data or code are updated, incorporating changes to websites or applications, testing modifications to software (\emph{e.g.}, R or Python libraries).
Action workflows can be found in GitHub's Marketplace (\url{https://github.com/marketplace?type=actions}) or, alternatively, in open user repositories.

@@

As an example, GitHub Organizations are particularly well-suited to host documents and projects within a laboratory, such as research compendia, codes of conduct, protocols, training documents, and other relevant documents that evolve collaboratively over time.
In this way, teams have full ownership of repositories within an organization, while ensuring that these materials stay accessible to the laboratory after people have moved on or when locally-stored data are lost.
This application extends to research centres, which may include several distinct projects that remain linked to institutions (\emph{e.g.}, the German Centre for Integrative Biodiversity Research (iDiv, \url{https://github.com/idiv-biodiversity})).
The team organizing the hackathon which inspired this article used a GitHub Organization (SORTEE-Github-Hackathon, \url{https://github.com/SORTEE-Github-Hackathon}) to centralize the project development, from meeting notes to, ultimately, this manuscript.
Organizations are also convenient for hosting learning materials, including lectures or workshops, such as the Québec Centre for Biodiversity Science R Workshop Series (QCBSRworkshops, \url{https://github.com/QCBSRworkshops}) or the University of Edinburgh's Coding Club (Coding Club, \url{https://github.com/ourcodingclub}), which may be continuously updated by an ever-evolving group of contributors.

@@

There have been many calls for researchers outside of the software development community to adopt GitHub to improve their collaborative research\textsuperscript{\protect\hyperlink{ref-10ghgV3S8}{3},\protect\hyperlink{ref-1Du6fzB8g}{39},\protect\hyperlink{ref-UsTxAq4f}{70}}.
These calls come in light of the continuous shift towards open-science and the increasing computational and data requirements in EEB.
Until now, resources and practical guidance on using GitHub within the EEB community have been dispersed in blog posts and video tutorials (\protect\hyperlink{tips}{Box 2}).
These resources have been useful for learning to use GitHub in our own work.
We expect that situating the main uses of GitHub in EEB into one medium, while adding on our personal perspectives, will be useful to the EEB community.

@@

First, real-time document editing is still best performed on other platforms (\emph{e.g.}, cloud-stored documents from Microsoft Word, Google Docs, hackMD (\url{https://hackmd.io/})).
Second, operations that are dependent on other software might not be easily achievable in GitHub, such as manipulating figures or tables.
Although creating tables and figures can be done through code, users may choose other software to collaboratively brainstorm figures and tables (\emph{e.g}., Google Slides, Google Sheets; but see GitHub Discussions).

@@

Lastly, when projects become highly collaborative they may have to pay for GitHub features such as branch protections, multiple reviewers of pull requests, and time in its automation tools.
Currently, GitHub offers Education Packs (\url{https://education.github.com/}) to students and academics, which extends some paid features to the free plan.
However, the acquisition of GitHub by Microsoft has raised concerns over the future of free plans, causing several biodiversity data managers to shift to alternative Open Source Git services (\emph{e.g.}, Bitbucket and GitLab).

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 0\tabcolsep) * \real{1.0000}}@{}}
\toprule
\begin{minipage}[b]{\linewidth}\raggedright
Ten tips for getting started in GitHub
\end{minipage} \\
\midrule
\endhead
1. \textbf{Check for existing solutions to your problem.} The GitHub Help webpage (\url{https://docs.github.com/en}) contains extensive and detailed documents with helpful screenshots. It is a good starting point for handling an issue, and has troubleshooting tips for specific problems. Alternatively, consider Tweeting your issue. There is a large community of GitHub users around the world who have likely faced analogous problems and may be able to provide quick solutions. Third, try to follow blogs (\emph{e.g.}, \url{https://github.blog}), Twitter accounts or YouTube channels that regularly post practical solutions about the most widely-used web platform for common GitHub issues. \\
2. \textbf{Consider taking free courses}, such as those from Software Carpentry\textsuperscript{\protect\hyperlink{ref-pjy75gHr}{73}}, and sharing these courses with your lab members or colleagues. \\
3. \textbf{Take advantage of GitHub as an asynchronous working tool for team-based projects.} See the repository for this paper (\url{https://github.com/SORTEE-Github-Hackathon/manuscript}) as an example of a collaboratively authored manuscript that used the GitHub Discussions, Issues, Pages, and Actions features. \\
4. \textbf{Use the interactive courses from the GitHub Skills page (\url{https://skills.github.com/})}, which allow you to learn GitHub basics through short projects and tasks with step-by-step guides. \\
5. \textbf{Learn markdown and use cheatsheets (\emph{e.g.}, \url{http://markdownguide.org/basic-syntax})} so you can write clear metadata README files for your repositories. \\
6. \textbf{Consult online resources.} The Jenny Bryan Universe of GitHub material, for example, provides a thorough and accessible introduction for a multitude of research-related uses for GitHub, and includes a book\textsuperscript{\protect\hyperlink{ref-ZvrOcg9w}{74}}, statistics course\textsuperscript{\protect\hyperlink{ref-6CMMeSeD}{75}} and academic article\textsuperscript{\protect\hyperlink{ref-RVetqmsg}{7}}. \\
7. \textbf{Do not be afraid of trial-and-error.} One of the best ways to learn GitHub is the ``trial-and-error'' method. Learning from your own mistakes can be a valuable way to master your GitHub abilities. In any case, if you make mistakes, GitHub allows you to revert any steps that you desire via version controlling. \\
8. \textbf{If you are an educator, include lectures on reproducibility and tools for creating reproducible workflows in the curricula.} Some graduate programs include coursework on course R Markdown and GitHub. Getting students started with these tools earlier will prevent the resistance that comes from working with a less reproducible workflow for a longer period of time (see example \url{https://github.com/rmcelreath/stat_rethinking_2022}). \\
9. \textbf{Try to begin committing with graphical user interfaces (GUI) instead of command line interfaces (CLI).} Examples of GUI are the GitHub Desktop (\url{https://desktop.github.com}), git-gui (\url{https://git-scm.com/docs/git-gui}), RStudio (\url{https://www.rstudio.com}), Visual Studio Code (\url{https://code.visualstudio.com}), Atom (\url{https://atom.io}), GitKraken (\url{https://www.gitkraken.com}). \\
10. \textbf{Get help deciphering GitHub Notifications.} Try using tools like Octobox (\url{https://octobox.io}) to disentangle and manage multiple notifications from distinct GitHub projects. \\
\bottomrule
\end{longtable}

@@

\hypertarget{ethics-declarations---competing-interests}{%
\subsection{Ethics declarations - Competing interests}\label{ethics-declarations---competing-interests}}

RCO was recently offered a position to work at GitHub (on July 15, 2022) and will start employment at GitHub on August 23, 2022.
Initial discussion of publishing this manuscript began in July 2021 and all work on the manuscript took place while RCO was an employee at Lawrence Berkeley National Laboratory.
